---
title: "Human Resources Employee Attrition"
author: "Attulaya"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

# Human Resources Employee Attrition

## Introduction

-   Employee attrition refers to the phenomenon of employees leaving an organization voluntarily or involuntarily.
-   High turnover rate is huge issues for organisations.
-   Predicting employee attrition can help organizations proactively identify and address factors that contribute to employee turnover.
-   High cost of hiring, training and time-consuming for recruitment.
-   Building a predictive model for employee attrition can provide valuable insights to optimize employee retention strategies.

## Problem Statement

-   Challenge in retaining current workforce

## Aim and Objectives

### Aim:

-   To develop a machine learning algorithm for accuracte and efficient detection of employee attrition.

### Objectives:

-   To investigate the key features affect employee attrition.

#### Let's look into what are all the installed libraries we have

```{r}
installed.packages()
```

#### Now we will look what are all the packages that are preloaded

```{r}
search()
```

#### let's load the packages

```{r}
library(pastecs)
library(DataExplorer)
```

#### I will load the data into a variable

```{r}
data<-read.csv("WA_Fn-UseC_-HR-Employee-Attrition.csv")
```

#### first we find out what is the class of the data

```{r}
class(data)
```

we found that the variable data is a data frame hence it is easy for to operate on it or else we would have to convert it into a data frame first and then perform the operations

#### lets take a look at the structure of data

```{r}
str(data)

```

#### now further moving lets get the summary of the current data before processing the data

```{r}
summary(data)
```

#### before manipulating the data let's have a look at the contents of the data frame by looking the head of the data frame and stats of the data

```{r}
head(data)
stat.desc(data)
dim(data)
```

### Data Pre-Processing

#### creating a Function Check if a value is blank

```{r}

is.blank <- function(x) {
  ifelse(is.na(x), TRUE, ifelse(is.character(x), trimws(x) == "", FALSE))
}
```

Replace blank values with NA in the entire data frame

```{r}
data[data == ""] <- NA
data[sapply(data, is.blank)] <- NA
```

#### now let's again look into the structure of data

```{r}
str(data)
```

#### Check duplicates based on id

```{r}
duplicates <- data[duplicated(data), ]
print(duplicates)
print(nrow(duplicates))
```

we find that there are no duplicate id's

#### Drop irrelevant feature

```{r}
data <- subset(data, select = -c(EmployeeCount, EmployeeNumber, StandardHours, Over18))
dim(data)


```

#### Lets's look at the Unique data

```{r}
data <- unique(data)
dim(data)
```

as there is no duplicate data and the number of unique rows is also the same after applying the unique function

### IN gender Replace "Male" with 1 and "Female" with 2

```{r}
data$Gender[data$Gender == "Male"] <- 1
data$Gender[data$Gender == "Female"] <- 2
```

### Convert the 'Gender' column back to a factor with correct levels

```{r}
data$Gender <- factor(data$Gender, levels = c("1", "2"))
```

### Lets get to know the Count of single and Married employees

```{r}
marital_single <- sum(data$MaritalStatus == "Single", na.rm = TRUE)
marital_married <-sum(data$MaritalStatus == "Married", na.rm = TRUE)
cat("Single:", marital_single, "\nMarried:", marital_married)

```


#### factorizing the values of attrition column
```{r}
data$Attrition <- factor(data$Attrition)
```


####loading the summarytools library
```{r}
library(summarytools)
```

#### Check data types
```{r}
print(class(data$MaritalStatus))
print(class(data$Attrition))
```

#### Check for missing values
```{r}

print(anyNA(data$MaritalStatus))
print(anyNA(data$Attrition))
```

#### Update summarytools package
```{r}

install.packages("summarytools", dependencies = TRUE)
```

#### creating a contingency table
```{r}

contingency_table <- ctable(data$MaritalStatus, data$Attrition, digits = 2)
```


#### looking into the contingency table
```{r}
head(contingency_table)
```

####  loading the dplyr library

```{r}
library(dplyr)
```


#### looking into the structure and dimensions of the data after the changes are made
```{r}
str(data)
```
```{r}
dim(data)
```

###### from the observation we see that the data is almost clean


##### To further clarify let's plot and see if there is any missing data
```{r}
colSums(is.na(data))
plot_missing(data)
```

## Replace missing values in numeric columns with mean using the ave function if it exists

```{r}
numeric_columns <- c("Age", "DailyRate", "HourlyRate", "MonthlyIncome", "MonthlyRate",
                     "NumCompaniesWorked", "PercentSalaryHike", "YearsAtCompany",
                     "YearsInCurrentRole", "YearsSinceLastPromotion", "YearsWithCurrManager",
                     "TrainingTimesLastYear")
for (col in numeric_columns) {
  data[[col]] <- ifelse(is.na(data[[col]]),
                      ave(data[[col]], FUN = function(x) mean(x, na.rm = TRUE)),
                      data[[col]])
}
```


##### Again let's have a look at the structure of data after making the changes
```{r}
str(data)
```
```{r}
df<-data
```


### Let's perform a bivariate analysis for each and every column  with respect to Attrition
```{r paged.print=FALSE}
library(ggplot2)
library(gridExtra)

# Select numerical variables for bivariate analysis
numerical_vars_bivariate <- c("Age", "DailyRate", "DistanceFromHome", "HourlyRate",
                              "MonthlyIncome", "MonthlyRate", "NumCompaniesWorked",
                              "PercentSalaryHike", "TotalWorkingYears", "TrainingTimesLastYear",
                              "YearsAtCompany", "YearsInCurrentRole", "YearsSinceLastPromotion",
                              "YearsWithCurrManager")

# Calculate the number of rows and columns for the plot layout
num_rows_num <- ceiling(length(numerical_vars_bivariate) / 4)
num_cols_num <- min(length(numerical_vars_bivariate), 4)

# Set the figure size
options(repr.plot.width = 40, repr.plot.height = 20 * num_rows_num)

# Create a list to store individual ggplot objects
plot_list_num <- vector("list", length = length(numerical_vars_bivariate))

# Create box plots for each numerical variable
for (i in seq_along(numerical_vars_bivariate)) {
  var <- numerical_vars_bivariate[i]
  p <- ggplot(data = data, aes(x = factor(Attrition), y = .data[[var]], fill = factor(Attrition))) +
    geom_boxplot() +
    labs(title = paste("Attrition vs.", var), y = var) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none")
  
  plot_list_num[[i]] <- p
}

# Display the plots using grid.arrange from the gridExtra package
library(gridExtra)
grid.arrange(grobs = plot_list_num, ncol = num_cols_num)
```

### lets create create some histograms to look the effects of other variables on attriation

#### - Educational Field
```{r}
library(repr)
options(repr.plot.width=10, repr.plot.height = 10) 
  ggplot(data=data)+
    geom_bar(aes(x=EducationField,color=Attrition,fill=Attrition))+
    ggtitle("Field of Education vs Attrition ") + 
    labs(x = "EducationField", y = "Count")

```
```{r}
library(dplyr)
data %>% 
    select(Attrition,Gender,EducationField) %>%
    filter (Attrition=="Yes") %>%
    group_by(EducationField,Gender) %>% 
    summarise(Count=n()) 
  
  attr_new <- data %>% 
    select(Attrition,Gender,EducationField) %>%
    filter (Attrition=="Yes") %>%
    group_by(EducationField,Gender) %>% 
    summarise(Count=n()) 

  ggplot(data=attr_new,aes(x=EducationField,y = Count,color=Gender,fill=Gender))+
    geom_bar(stat="identity", position=position_dodge())+
    ggtitle("Field of Education vs Gender ")+
  labs(x = "EducationField", y = "Count")
```


#### - Monthly Income
```{r}
data %>% 
    summarise(Median = median(MonthlyIncome), 
              Mean = mean(MonthlyIncome),
              Max = max(MonthlyIncome), 
              Min = min(MonthlyIncome))
```
```{r}
ggplot(data, aes(x=Attrition, y=MonthlyIncome, color=Gender, fill=Gender)) +
    geom_boxplot()
```
```{r}
ggplot(data, aes(x=MonthlyIncome,color=Attrition,fill=Attrition)) +
    geom_histogram(position="identity", alpha=0.7)+
    ggtitle("Distribution of Monthly Income ")+
    labs(x = "Monthly Income", y = "Count")
```


#### Relation between monthly income and overtime 
```{r}
ggplot(data, aes(x=OverTime, y=MonthlyIncome, color=Gender, fill=Gender)) +
    geom_boxplot()
```


#### Relation between buisness travell and and years at company
```{r}
library(ggplot2)
ggplot(data = data) + geom_boxplot(aes(x = factor(BusinessTravel), y = YearsAtCompany)) +
xlab('YearsAtCompany') + labs(title = 'Distribution of hours by BusinessTravel')
```


#### Dempartment and years at company
```{r}
ggplot(data =data) + geom_boxplot(aes(x = factor(Department), y = YearsAtCompany))
```


#### Years since last promotion and years at company
```{r}
ggplot(data =data) + geom_point(aes(x = YearsSinceLastPromotion, y = YearsAtCompany))
```


### Normalizing the data to create a correlation matrix
```{r}
library(caret)
library(dplyr)
ds1 <- data %>% select (Department, EducationField, MaritalStatus, JobRole)
ds2 <- data %>% select (-Department, -EducationField, -MaritalStatus, -JobRole)

dmy <- dummyVars("~.", data=ds1)
onehot <- data.frame(predict(dmy, newdata=ds1))
head(onehot)

```
```{r}
final <- cbind(onehot,ds2)
```

##### Have a look at structure of data
```{r}
str(final)
```


###Normalization
```{r}
library(caret)

# Exclude non-numeric columns and the target variable
numeric_cols <- final[, sapply(final, is.numeric) & !colnames(final) %in% "Attrition"]

# Function to apply Min-Max scaling on the dataset
min_max_scaling <- function(x) {
  return((x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)))
}

# Apply Min-Max scaling on the numerical variables
normalized_data <- as.data.frame(lapply(numeric_cols, min_max_scaling))

# Combine the normalized numerical variables with the target variable (Attrition)
final_normalized <- cbind.data.frame(normalized_data, Attrition = final$Attrition)

# Print the structure of the final normalized dataset
str(final_normalized)

```


###### Renaming
```{r}
# Rename the 'final$Attrition' column to 'Attrition'
colnames(final_normalized)[colnames(final_normalized) == "final$Attrition"] <- "Attrition"
final_normalized$Attrition <- as.numeric(final_normalized$Attrition)
```


#### Plotting of correlation matrix
```{r}
library(reshape2)
correlation_matrix <- cor(final_normalized)
correlation_matrix
ggplot(data = melt(correlation_matrix), aes(x = Var1, y = Var2, fill = value, label = round(value, ))) +
  geom_tile() +
  geom_text(color = "black", size = 2) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Correlation Matrix Heatmap")
```


#### Calculating the average age
```{r}
avg.age <- data %>% select(Gender, Age) %>% group_by(Gender) %>% summarize(avg=mean(Age))

avg.age
```



#### Ploting the density distribution graph for age based on genders
```{r}
library(cowplot)
library(dplyr)
library(ggplot2)

options(repr.plot.width=8, repr.plot.height=6) 

dat_text <- data.frame(
  label = c("Mean = 37.33 \n Years Old", "Mean = 36.65 \n Years Old"),
  Gender = c("Female", "Male")
)

# Example data
set.seed(123)
data <- data.frame(
  Gender = sample(c("Male", "Female"), 100, replace = TRUE),
  Age = rnorm(100, mean = 37, sd = 5)
)

gender.dist <- data %>% 
  select(Gender, Age) %>% 
  filter(Gender == 'Male' | Gender == "Female") %>% 
  filter(!is.na(Age)) %>% 
  group_by(Gender) %>% 
  ggplot(aes(x=Age)) + 
  geom_density(aes(fill=Gender), alpha=0.8, show.legend=FALSE) + 
  facet_wrap(~Gender) + 
  theme_minimal() + 
  geom_vline(aes(xintercept=mean(Age)),
             color="red", linetype="dashed", linewidth=1) + 
  labs(title="Age Distribution") + 
  theme(plot.title=element_text(hjust=0.5)) + 
  scale_fill_manual(values=c("#F781F3", "#819FF7")) + 
  geom_text(
    data = dat_text,
    mapping = aes(x = 45, y = 0.03, label = label),
    hjust = -0.1,
    vjust = -1
  )

overall.dist <- data %>% 
  select(Gender, Age) %>% 
  filter(!is.na(Age)) %>% 
  ggplot(data=data, mapping=aes(x=Age)) + 
  geom_density(color="darkblue", fill="lightblue") + 
  geom_vline(aes(xintercept=mean(Age)),
             color="red", linetype="dashed", linewidth=1) +  
  theme_minimal() + 
  labs(x="Overall Age") + 
  annotate("text", label = "Mean = 36.92 Years Old", x = 50, y = 0.03, color = "black")

plot_grid(gender.dist, overall.dist, nrow=2)


```


- The companies overall age groups almost follows the bell-curve stating that the company has well diversified employees in terms of age groups
- The age of women who's age is more than the average age of women are more in number 
- The age of men who's age is lesser than the average age of men are more in number 


#### Gender and income relation 
```{r}
library(ggplot2)
p <- ggplot(df, aes(x=Gender, y= MonthlyIncome, color=Gender, fill=Gender)) + geom_boxplot() + 
scale_fill_manual(values=c("#F5A9F2", "#5882FA")) + scale_color_manual(values=c("#FE2EF7", "#5858FA")) +
coord_flip() + labs(title="Are there any Gender Disparities in Income?")

p
```
from the above graph we can see that there is a slight disparity in the income 


#### Attrition in research and development department
```{r}
library(dplyr)
r.d <- df %>% select(Department, WorkLifeBalance, Attrition) %>% 
filter(Department == "Research & Development" & WorkLifeBalance == 1 | WorkLifeBalance == 2) %>%
group_by(Attrition) %>% summarize(count=n())
r.d
```





